GPU-Beschleunigung der Material Point Method

### 2018-11-12 mscthesis.tex
polar svd

\cite{MPM:GPU}
%look again for MPM/PIC gpu impl
%SPH, FEM methods
\section{Notation}
Multi-component types are generally printed in bold letters $\boldsymbol{A}$, $\boldsymbol{a}$. Vectors furthermore use lower-case letters $\boldsymbol{v}$, while Matrices use upper-case letters $\boldsymbol{M}$. A variable in the Lagrangian form has a subscript 0 $ _0{\boldsymbol{x}}$, the Eulerian description will be written with an 'overline' $_t\boldsymbol{x}$ if distinction is needed. Occasionally notation may be omitted where it is apparent from context.
\subsection{Einstein-Notation}
Used throughout this thesis is at instances the Einstein-Notation when using tensors and vectors. Let $\boldsymbol{a}$ and $\boldsymbol{b}$ be vectors of dimension $n$ and  $\boldsymbol{A,B,D}$ $ { m \times n }$ tensors. Repeated indices that are also defined on the variable imply component-wise operations.
$a _ { i } b _ { j }$ is multiplying component $i$ with component $j$ of vectors $\boldsymbol{a}$ and $\boldsymbol{b}$. Vector and tensor/matrix addition thus become:
\begin{equation} \label{EINSTEIN:ADD}
  c _ { i } = a _ { i } + b _ { i } \quad \text { and } \quad D _ { i j } = A _ { i j } + B _ { i j }
\end{equation}
Repeated indices that are not otherwise defined however imply summation on that index. The vector dot product becomes:
\begin{equation}
  a _ { i } b _ { i } \equiv \sum _ { i = 1 } ^ { n } a _ { i } b _ { i }
\end{equation}
Following this notation the Frobenius inner product between two second-order tensors is (also called Frobenius scalar product):
\begin{equation} \label{EINSTEIN:FROBENIUS1}
  \boldsymbol { A } : \boldsymbol { B } \equiv A _ { i j } B _ { i j } \equiv \sum _ { i = 1 } ^ { n } \sum _ { j = 1 } ^ { m } A _ { ij } B _ { ij }
\end{equation}
The Frobenius inner product between a ${r \times s \times m \times n }$ fourth-order tensor $\boldsymbol{C}$  and second-order tensor creates a second order tensor combining the ideas of \ref{EINSTEIN:ADD} and \ref{EINSTEIN:FROBENIUS1}:
\begin{equation}
  \boldsymbol { A } = A_ {ij} =  \boldsymbol {C} : \boldsymbol { B } = C _ {i j k l} B _ { k l  } = \sum _ { k = 1 } ^ { m } \sum _ { l = 1 } ^ { n } C _ { ijkl } B _ { kl }
\end{equation}
Matrix-Vector and Matrix-Matrix multiplication can be expressed this way as:
\begin{equation}
  b_i = A_{i j} a_j \quad D _ { i j } = A _ { i k } B _ { k j }
\end{equation}
The definition of the Kronecker Delta is as follows:
\begin{equation}
  \delta _ { i j } = \left\{
    \begin{array} { l l l }
      { 1 } & {\text{if}} &  {i = j}  \\
      { 0 } & {\text{if}} &  { i \neq j }
  \end{array} \right.
\end{equation}
The Kronecker Delta as an operator is most efficiently described as a substitution:
\begin{equation}
  a _ { i } \delta _ { i j } = a _ { j }
\end{equation}
I.e. this component is only evaluated if $i = j$. In this case it becomes equivalent to the identity matrix $\boldsymbol{I}$.
\cite{MCGINTY:CONTINUUM}
\section{Basics}
\subsection{Reference and current configuration}
Particles (or material points) in continuum mechanics are not what classically might be thought of as a particle. Rather the continuum assumption holds: each particle represents a continuous piece of material, s.t. a microscopic view does not need to be adopted. A particular body is composed of a set of particles and can adapt different configurations due to changes in shape. These changes are cause by external or internal effects (forces etc.) on it and deform the body over time. \cite{MIT:CONTINUUM_MECHANICS} \cite{MPM:COURSE}

When modelling solids changes of quantities from one configuration (reference) $_0\boldsymbol{x}$ to another configuration (current) $_t\boldsymbol{x}$ need to be measured. In the material point method the reference configuration $^0\boldsymbol{x}$ is just the initial configuration of the body (at time $t=0$). This is similar to the total Lagrangian formulation in finite element methods \cite{bathe2006finite}.
Let $\Omega ^ { 0 } , \Omega ^ { t } \subset \mathbb { R } ^ {  { 3 } }$ be the set of (material) points in the initial and current configuration respectively.  Then one may define a function or mapping ${^t_0}\phi ( \cdot , t ) : \Omega ^ { 0 } \rightarrow \Omega ^ { t }$, which relates the initial configuration to the current configuration. If we assume for simplicity that $_0\boldsymbol{x}$ and $_t\boldsymbol{{x}}$ describe the position of the particle in their respective configurations, this mapping becomes the deformation of the body from the reference configuration $_0\boldsymbol{x}$:
\begin{equation}
  _t\boldsymbol{x} =  {^t_0}\phi ( _0\boldsymbol{x} , \boldsymbol { t } )
\end{equation}
If for instance the body consisting of each material point $_t\boldsymbol{x}$ moves with velocity $\boldsymbol{v}$ and rotation $\boldsymbol{R}(t)$ , we can defines this mapping to be:
\begin{equation}
  _t\boldsymbol{{x}} = {^t_0}\phi ( _0\boldsymbol{x} , \boldsymbol { t } ) = \boldsymbol{R}(t)_0\boldsymbol{x}+\boldsymbol{v}t
\end{equation}
The velocity of a material point in $_0\boldsymbol{x}$ can be defined using this mapping
\begin{equation}\label{eq:velocity}
  _0\boldsymbol{v}(_0\boldsymbol{x},t) = \frac{\partial {^t_0}\phi}{\partial t}(_0\boldsymbol{x},t)
\end{equation}
and similarly the acceleration is defined
\begin{equation}
  _0\boldsymbol{a}(_0\boldsymbol{x},t) = \frac{\partial^2 {^t_0}\phi}{\partial t^2}(_0\boldsymbol{x},t) = \frac{\partial _0\boldsymbol{v}}{\partial t}(_0\boldsymbol{x},t).
\end{equation}
It is helpful to abstract away from the reference configuration and think of it as being defined in a different fixed material space. Physically this has the impact of moving with the particle in world space. Commonly known as the Lagrangian form. It is often easier in continuum mechanics to start with a Lagrangian description and switch to a Eulerian one if needed. The Eulerian description is static: Variables of particles that are moving by are measured while staying in a fixed position.
These descriptions are different but they will yield the same measurements when related correctly. These relations for some particle quantity $f$ are called the (Lagrangian) pull back
\begin{equation}
  _0f(_0\boldsymbol{x},t)= {_tf}({^t_0}\phi(_0\boldsymbol{x},t),t)
\end{equation}
and the (Eulerian) push forward
\begin{equation}\label{eq:push_forward}
  _tf(_t\boldsymbol{x},t) =  {_0f}({^t_0}\phi^{-1}(_t\boldsymbol{x},t),t) = {_0f}({_t^0}\phi(_t\boldsymbol{x},t),t)
\end{equation}
with definitions over their respective spaces $_tf ( \cdot , t ) : \Omega ^ {  { t } } \rightarrow \mathbb { R }$, $_0f ( \cdot , t ) : \Omega ^ {   { 0 } } \rightarrow \mathbb { R }$.  To enable the operator ${^t_0}\phi$ to be homeomorphic, s.t. an inverse ${^t_0}\phi^{-1} = {_t^0}\phi$ is defined, it is assumed that no two particles will ever occupy the same space at the same time.

@@

The area change is given by Nanson's Formula. Where $\boldsymbol{A}$ points in the direction of the normal of the area:

@@

A surface intergral may then be transformed to current configuration by

@@

where $_0h = h(_0\boldsymbol{x},\cdot)$ is the pull back of $_th = h(_t\boldsymbol{x},\cdot)$.

@@

\cite{SVD:3x3}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%                         Constitutive Models                         %
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

@@

The energy produced by equation \ref{eq:linear_elas} using the small strain tensor is not rotationally invariant, w.r.t. to $\boldsymbol{F}_E$: $\psi(\boldsymbol{\epsilon}(\boldsymbol{R}_{0} \boldsymbol{F}_E)) \neq \psi(\boldsymbol{\epsilon}(\boldsymbol{F}_E))$. Rigid body motions however don't result in deformation and consequently don't need to be recovered from. So energy should not change. Given the polar decomposition $\boldsymbol{F}_E = \boldsymbol{R}_E\boldsymbol{S}_E$, an alternate strain measure may be defined as:

>>>>

\cite{MPM:GPU}
%look again for MPM/PIC gpu impl
%SPH, FEM methods
\section{Notation}
Multi-component types are generally printed in bold letters $\boldsymbol{A}$, $\boldsymbol{a}$. Vectors furthermore use lower-case letters $\boldsymbol{v}$, while Matrices use upper-case letters $\boldsymbol{M}$. A variable in the Lagrangian form has a subscript 0 $ _0{\boldsymbol{x}}$, the Eulerian description will be written with an 'overline' $_t\boldsymbol{x}$ if distinction is needed. Occasionally notation may be omitted where it is apparent from context.
\subsection{Einstein-Notation}
Used throughout this thesis is at instances the Einstein-Notation when using tensors and vectors. Let $\boldsymbol{a}$ and $\boldsymbol{b}$ be vectors of dimension $n$ and  $\boldsymbol{A,B,D}$ $ { m \times n }$ tensors. Repeated indices that are also defined on the variable imply component-wise operations.
$a _ { i } b _ { j }$ is multiplying component $i$ with component $j$ of vectors $\boldsymbol{a}$ and $\boldsymbol{b}$. Vector and tensor/matrix addition thus become:
\begin{equation} \label{EINSTEIN:ADD}
  c _ { i } = a _ { i } + b _ { i } \quad \text { and } \quad D _ { i j } = A _ { i j } + B _ { i j }
\end{equation}
A transpose operator swaps the indices:
\begin{equation}
  D^T _ { i j }  =D_{ji}
\end{equation}
Repeated indices that are not otherwise defined however imply summation on that index. The vector dot product becomes:
\begin{equation}
  a _ { i } b _ { i } \equiv \sum _ { i = 1 } ^ { n } a _ { i } b _ { i }
\end{equation}
Following this notation the Frobenius inner product between two second-order tensors is (also called Frobenius scalar product):
\begin{equation} \label{EINSTEIN:FROBENIUS1}
  \boldsymbol { A } : \boldsymbol { B } \equiv A _ { i j } B _ { i j } \equiv \sum _ { i = 1 } ^ { n } \sum _ { j = 1 } ^ { m } A _ { ij } B _ { ij }
\end{equation}
The Frobenius inner product between a ${r \times s \times m \times n }$ fourth-order tensor $\boldsymbol{C}$  and second-order tensor creates a second order tensor combining the ideas of \ref{EINSTEIN:ADD} and \ref{EINSTEIN:FROBENIUS1}:
\begin{equation}
  \boldsymbol { A } = A_ {ij} =  \boldsymbol {C} : \boldsymbol { B } = C _ {i j k l} B _ { k l  } = \sum _ { k = 1 } ^ { m } \sum _ { l = 1 } ^ { n } C _ { ijkl } B _ { kl }
\end{equation}
Matrix-Vector and Matrix-Matrix multiplication can be expressed this way as:
\begin{equation}
  b_i = A_{i j} a_j \quad D _ { i j } = A _ { i k } B _ { k j }
\end{equation}
The definition of the Kronecker Delta is as follows:
\begin{equation}
  \delta _ { i j } = \left\{
    \begin{array} { l l l }
      { 1 } & {\text{if}} &  {i = j}  \\
      { 0 } & {\text{if}} &  { i \neq j }
  \end{array} \right.
\end{equation}
The Kronecker Delta as an operator is most efficiently described as a substitution:
\begin{equation}
  a _ { i } \delta _ { i j } = a _ { j }
\end{equation}
I.e. this component is only evaluated if $i = j$. In this case it becomes equivalent to the identity matrix $\boldsymbol{I}$. The Kronecker Delta comes up in differentiating a variable by itself with different indexing:
\begin{equation}
  \frac{\partial{x_0}}{\partial{x_1}}=0, \frac{\partial{x_0}}{\partial{x_0}}=1  \Rightarrow \frac{\partial{x_i}}{\partial{x_j}} = \delta_{ij}
\end{equation}
\cite{MCGINTY:CONTINUUM}
\section{Basics}
\subsection{Reference and current configuration}
Particles (or material points) in continuum mechanics are not what classically might be thought of as a particle. Rather the continuum assumption holds: each particle represents a continuous piece of material, s.t. a microscopic view does not need to be adopted. A particular body is composed of a set of particles and can adapt different configurations due to changes in shape. These changes are cause by external or internal effects (forces etc.) on it and deform the body over time. \cite{MIT:CONTINUUM_MECHANICS} \cite{MPM:COURSE}

When modelling solids changes of quantities from one configuration (reference) $_0\boldsymbol{x}$ to another configuration (current) $_t\boldsymbol{x}$ need to be measured. In the material point method the reference configuration $^0\boldsymbol{x}$ is just the initial configuration of the body (at time $t=0$). This is similar to the total Lagrangian formulation in finite element methods \cite{bathe2006finite}.
Let $\Omega ^ { 0 } , \Omega ^ { t } \subset \mathbb { R } ^ {  { 3 } }$ be the set of (material) points in the initial and current configuration respectively.  Then one may define a function or mapping ${^t_0}\phi ( \cdot , t ) : \Omega ^ { 0 } \rightarrow \Omega ^ { t }$, which relates the initial configuration to the current configuration. If we assume for simplicity that $_0\boldsymbol{x}$ and $_t\boldsymbol{{x}}$ describe the position of the particle in their respective configurations, this mapping becomes the deformation of the body from the reference configuration $_0\boldsymbol{x}$:
\begin{equation}
  _t\boldsymbol{x} =  {^t_0}\phi ( _0\boldsymbol{x} , \boldsymbol { t } )
\end{equation}
If for instance the body consisting of each material point $_t\boldsymbol{x}$ moves with velocity $\boldsymbol{v}$ and rotation $\boldsymbol{R}(t)$ , we can defines this mapping to be:
\begin{equation}\label{eq:rigid}
  _t\boldsymbol{{x}} = {^t_0}\phi ( _0\boldsymbol{x} , \boldsymbol { t } ) = \boldsymbol{R}(t)_0\boldsymbol{x}+\boldsymbol{v}t
\end{equation}
The velocity of a material point in $_0\boldsymbol{x}$ can be defined using this mapping
\begin{equation}\label{eq:velocity}
  _0\boldsymbol{v}(_0\boldsymbol{x},t) = \frac{\partial {^t_0}\phi}{\partial t}(_0\boldsymbol{x},t)
\end{equation}
and similarly the acceleration is defined
\begin{equation}
  _0\boldsymbol{a}(_0\boldsymbol{x},t) = \frac{\partial^2 {^t_0}\phi}{\partial t^2}(_0\boldsymbol{x},t) = \frac{\partial _0\boldsymbol{v}}{\partial t}(_0\boldsymbol{x},t).
\end{equation}
It is helpful to abstract away from the reference configuration and think of it as being defined in a different fixed material space. Physically this has the impact of moving with the particle in world space. Commonly known as the Lagrangian form. It is often easier in continuum mechanics to start with a Lagrangian description and switch to a Eulerian one if needed. The Eulerian description is static: Variables of particles that are moving by are measured while staying in a fixed position.
These descriptions are different but they will yield the same measurements when related correctly. These relations for some particle quantity $f$ are called the (Lagrangian) pull back
\begin{equation}
  _0f(_0\boldsymbol{x},t)= {_tf}({^t_0}\phi(_0\boldsymbol{x},t),t)
\end{equation}
and the (Eulerian) push forward
\begin{equation}\label{eq:push_forward}
  _tf(_t\boldsymbol{x},t) =  {_0f}({^t_0}\phi^{-1}(_t\boldsymbol{x},t),t) = {_0f}({_t^0}\phi(_t\boldsymbol{x},t),t)
\end{equation}
with definitions over their respective spaces $_tf ( \cdot , t ) : \Omega ^ {  { t } } \rightarrow \mathbb { R }$, $_0f ( \cdot , t ) : \Omega ^ {   { 0 } } \rightarrow \mathbb { R }$.  To enable the operator ${^t_0}\phi$ to be homeomorphic, s.t. an inverse ${^t_0}\phi^{-1} = {_t^0}\phi$ is defined, it is assumed that no two particles will ever occupy the same space at the same time.

@@

The area change is given by Nanson's Formula. Where $d\boldsymbol{A}$ is a vector pointing in the direction of the normal of the area:

@@

A surface integral may then be transformed to reference configuration by

@@

where $_0\boldsymbol{h} = \boldsymbol{h}(_0\boldsymbol{x},\cdot)$ is the pull back of $_t\boldsymbol{h} = \boldsymbol{h}(_t\boldsymbol{x},\cdot)$.

@@

The target is to define strain measures in terms of the deformation gradient: $\boldsymbol{\epsilon}(\boldsymbol{F})$. In equation \ref{eq:rigid} a rigid body movement was introduced. A problem arises when calculating the deformation gradient of this equation.
Let $b_i(t) = v_it$ be more generally a translation:
  ^t_0F_{ij} = \frac{\partial{_tx_i}}{\partial_0x_j} = \frac{\partial (R_{ik}(t)_0x_k+b_i(t))}{\partial _0x_j} = R_{ik}(t) \delta_{kj} = R_{ij}(t)
As can be seen the deformation gradient contains a rigid rotation. For strain measures this is not beneficial as an assumption of the stiffness tensor requires no net-rotation. I.e. the deformation gradient has two components a constant rotation and the actual distortion or strain. There is two ways to deal with this:
  \item  Use a strain measure that cancels out the rotation. An example for this would be the Green-strain with quadratic components:
    \begin{equation}\label{eq:green}
      E _ { i j } = \frac { 1 } { 2 } \left( F _ { k i } F _ { k j } - \delta _ { i j } \right).
  \item \label{it:polar} Polar Decompose the deformation gradient in its rotational $\boldsymbol{R}$ and (symmetric) distortion $\boldsymbol{U}$ part $\boldsymbol{F}=\boldsymbol{U}\boldsymbol{R}$.
That equation \ref{eq:green} cancels out the rotational part can be shown by item \ref{it:polar}
$$ \frac { 1 } { 2 } \left( F _ { k i } F _ { k j } - \delta _ { i j } \right) = \frac { 1 } { 2 } \left( U _ {m i}R_{k m} R _ { k n} U_{n j} - \delta _ { i j } \right)$$
$$= \frac { 1 } { 2 } \left( U _ {m i} \delta_{mn} U_{n j} - \delta _ { i j } \right) = \frac { 1 } { 2 } \left( U _ {i m} U_{m j} - \delta _ { i j } \right)$$
$$= \frac{1}{2} \left(\boldsymbol{U}^2 - I \right).$$

\cite{SVD:3x3}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%                         Constitutive Models                         %
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

@@

The energy produced by equation \ref{eq:linear_elas} using the small strain tensor is not rotationally invariant, w.r.t. to $\boldsymbol{F}_E$: $\psi(\boldsymbol{\epsilon}(\boldsymbol{R}_{0} \boldsymbol{F}_E)) \neq \psi(\boldsymbol{\epsilon}(\boldsymbol{F}_E))$. Rigid body motions however don't result in strain and consequently don't need to be recovered from. So energy should not change. Given the polar decomposition $\boldsymbol{F}_E = \boldsymbol{R}_E\boldsymbol{S}_E$, an alternate strain measure may be defined as:
